---
title: "midterm-simulation-project"
author: "Parker Hicks"
date: "2023-10-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
```

## Set simulated data
```{r}
set.seed(1234)
# PRS scores tend to be normally distributed
EUR_scores <- rnorm(n=10000, mean=0, sd=2)
AFR_scores <- rnorm(n=5000, mean=-1, sd=2)
```

### Merge data
```{r}
EUR_df <- data.frame(scores=EUR_scores, pop="EUR", stringsAsFactors=TRUE)
AFR_df <- data.frame(scores=AFR_scores, pop="AFR", stringsAsFactors=TRUE)

merged_pops <- rbind(EUR_df, AFR_df)
```

## Plot population scores
```{r}
population_scores_density <- ggplot(data=merged_pops, aes(x=scores, fill=pop)) +
  geom_density(alpha=0.5) +
  xlab("PRS")

population_scores_histogram <- ggplot(data=merged_pops, aes(x=scores, fill=pop)) +
  geom_histogram(binwidth=1, alpha=0.5) +
  xlab("PRS") +
  ggtitle("Absolute counts of PRS scores across simulated EUR and AFR populations")

population_scores_density
ggsave(
  "./imgs/population_scores_density.png",
  plot = last_plot(),
  dpi = 500,
)
population_scores_histogram
```
PRS scores typically follow a normal distribution 


RQ: How does boostrapping the mean difference from 1 sample compare to the computing the mean difference across many independent samples?

```{r}
set.seed(1234)

B <- 10^4
mean_differences <- numeric(B)
for (i in 1:B) {
  # Sample populations from a normal distribution
  eur_pop <- rnorm(10000, mean=0, sd=2)
  afr_pop <- rnorm(5000, mean=-1, sd=2)

  # Get the top 15th percentiles from both distributions
  eur_15th <- quantile(eur_pop, probs=c(0.85))
  afr_15th <- quantile(afr_pop, probs=c(0.85))
  
  # Compute mean difference statistic
  mean_diff <- mean(eur_pop[eur_pop > eur_15th]) - 
    mean(afr_pop[afr_pop > afr_15th])
  mean_differences[i] <- mean_diff
}

difference_samples_df <- data.frame(pop=1:B, pop_differences=mean_differences)
```

```{r}
difference_samples_hist <- ggplot(data=difference_samples_df, aes(x=pop_differences)) +
            geom_histogram(binwidth = 0.05) +
            xlab("Difference of upper 15th percentile means")

difference_samples_qq <- ggplot(data=difference_samples_df, aes(sample=pop_differences)) +
          stat_qq() +
          stat_qq_line()

difference_samples_hist
ggsave(
  "./imgs/difference_samples_hist.png",
  plot = last_plot(),
  dpi = 500,
)
difference_samples_qq
```

## Subset populations by top 15th percentile
```{r}
# Subset population PRS scores
EUR_15th <- quantile(EUR_df$scores, probs=c(0.85))
AFR_15th <- quantile(AFR_df$scores, probs=c(0.85))

EUR_topPercentile <- subset(EUR_df, scores >= EUR_15th)
AFR_topPercentile <- subset(AFR_df, scores >= AFR_15th)
```

## Perform difference boostrap
```{r}
set.seed(1234)
boostrap_difference <- function(sample1, sample2, iterations) {
  boostrap_distribution <- numeric(iterations)
  for (i in 1:iterations) {
  x <- sample(sample1, length(sample1), replace=TRUE)
  y <- sample(sample2, length(sample2), replace=TRUE)
  boostrap_distribution[i] <- mean(x) - mean(y)
  }
  
  return(boostrap_distribution)
}

B <- 10^4
prs_difference_boot <- boostrap_difference(sample1=EUR_topPercentile$scores,
                                      sample2=AFR_topPercentile$scores,
                                      iterations=B
                                      )
difference_boot_df <- data.frame(pop=1:B, pop_differences=prs_difference_boot)
```


## Plot boostraped statistic
```{r}
difference_boot_hist <- ggplot(data=difference_boot_df, aes(x=pop_differences)) +
            geom_histogram(binwidth = 0.05) +
            xlab("Difference of upper 15th percentile means")

difference_boot_qq <- ggplot(data=difference_boot_df, aes(sample=pop_differences)) +
          stat_qq() +
          stat_qq_line()

difference_boot_hist
ggsave(
  "./imgs/difference_boot_hist.png",
  plot = last_plot(),
  dpi = 500,
)
difference_boot_qq
```

```{r}
samples_mean <- mean(difference_samples_df$pop_differences)
boot_mean <- mean(difference_boot_df$pop_differences)

samples_se <- sd(difference_samples_df$pop_differences)
boot_se <- sd(difference_boot_df$pop_differences)

samples_var <- sd(difference_samples_df$pop_differences)^2
boot_var <- sd(difference_boot_df$pop_differences)^2

samples_CI <- mean(difference_samples_df$pop_differences) + quantile(difference_samples_df$pop_differences, 0.05 / 2) * 2 / sqrt(length(difference_samples_df$pop_differences))
boot_CI <- quantile( difference_boot_df$pop_differences, c(0.025,0.975))

samples_mean
boot_mean

samples_se
boot_se

samples_var
boot_var

samples_CI
boot_CI
```

## Compute the bias of the boostrap estimator
```{r}
bias <- mean(difference_boot_df$pop_differences) - mean(difference_samples_df$pop_differences)
bias
```